---
title: "Sex-dependent response calling in Thyroptera tricolor"
subtitle: "Statistical analysis"
author: <a href="http://marceloarayasalas.weebly.com/">Marcelo Araya-Salas</a>
 &nbsp; 
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: no
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
fontsize: 12pt 
editor_options: 
  chunk_output_type: console
---


```{r packages, message=FALSE, warning = FALSE, echo = TRUE, eval = TRUE, include = FALSE}

# unload all non-based packages
out <- sapply(paste('package:', names(sessionInfo()$otherPkgs), sep = ""), function(x) try(detach(x, unload = FALSE, character.only = TRUE), silent = T))

## add 'developer/' to packages to be installed from github
x <- c("ggplot2", "readxl", "viridis", "MCMCglmm", "ranger", "warbleR", "bioacoustics", "Sim.DiffProc")

aa <- lapply(x, function(y) {
  
  # get pakage name
  pkg <- strsplit(y, "/")[[1]]
  pkg <- pkg[length(pkg)]
  
  # check if installed, if not then install 
  if (!pkg %in% installed.packages()[,"Package"])  {

      if (grepl("/", y))  devtools::install_github(y, force = TRUE) else
    install.packages(y) 
    }

  # load package
  try(require(pkg, character.only = T), silent = T)
})

```

```{r functions, eval = TRUE, echo = TRUE}

# path to project directory
knitr::opts_knit$set(root.dir = normalizePath(".."))

# fig quality
knitr::opts_chunk$set(dpi = 50, fig.width = 12) 

num.trees <- 10000
```

## Exploratory graphs
```{r read data, fig.height = 6, eval = TRUE}

# read data
dat <- read_excel("./data/Datos de respuestas_cap2.xlsx")

# convert to regular data frame
dat <- as.data.frame(dat)

# create new variable abou
dat$est_repr[dat$estado_repr == 0] <- "Inactivo"
dat$est_repr[dat$estado_repr == 1] <- "Activo"

# aggregate total number of calls
agg_dat <- aggregate(n_llamadas ~ ID + sexo_consulta + sexo_respuesta + est_repr, data = dat, FUN = sum)

# plot
ggplot(agg_dat, aes(fill = sexo_consulta, y = n_llamadas, x = sexo_respuesta)) +
geom_boxplot() +
  scale_fill_viridis_d(alpha = 0.7, begin = 0.4) +
theme_classic(base_size = 24) +
  labs(x = "Sexo respuesta", y = "Numero de llamadas")+
facet_wrap(~ est_repr) + ggtitle("Total de llamadas")


# create binary variable for calling
dat$n_llam_bin <- ifelse(dat$n_llamadas > 0, 1, 0)

# aggregate
agg_dat2 <- aggregate(n_llam_bin ~ ID + sexo_consulta + sexo_respuesta + est_repr, data = dat, FUN = sum)

#plot 
ggplot(agg_dat2, aes(fill = sexo_consulta, y = n_llam_bin, x = sexo_respuesta)) +
  geom_boxplot() +
  scale_fill_viridis_d(alpha = 0.7, begin = 0.4) +
  theme_classic(base_size = 24) +
  labs(x = "Sexo respuesta", y = "Cuenta de experimentos con respuesta")+
facet_wrap(~ est_repr) + ggtitle("Experimentos con respuesta")

```


## Multinomial mixed effect models (MCMCglmm)

- Including three-way interaction between 'sex response', 'sex inquiry' and 'reproductive stage', with individual ID as random effect:

```{r MCMCglmm, eval = FALSE}

# define parmeters for MCMCglmm models
itrns <- 300000
burnin <- 3000
thin <- 1000

# prior for effect models
pr <- list(B = list(mu = rep(0, 8), V = diag(8) * (1 + pi^2/3)), R = list(V = 1, fix = 1),  G = list(G1 = list(V = 1, fix = 1)))

# run model
md <- MCMCglmm(n_llam_bin ~ sexo_respuesta:est_repr:sexo_consulta  - 1, random = ~ ID, data = dat, family = "categorical", prior = pr, verbose = FALSE, nitt = itrns, start = list(QUASI = FALSE), burnin = burnin, thin = thin)

saveRDS(md, "./output/MCMC_model_response_by_sex_test.RDS")

```

# Caculating p-values:

The output contains the posterior distribution of the parameter estimates. These parameter distributions can be used to test specific hypothesis about differences between sexes/stages/inquiry sexes. Column names in `md$Sol` (solutions) refer to the combination of levels from the 3 interacting variables:

```{r read MCMC results, fig.height = 14, eval = FALSE}

md <- readRDS("./output/MCMC_model_response_by_sex_test.RDS")


# simplify names
colnames(md$Sol) <- gsub("sexo_respuesta|est_repr", "", colnames(md$Sol))

colnames(md$Sol)

# stack posteriors
Y <- stack(as.data.frame(md$Sol))

# plot posteriors
ggplot(Y, aes(x=values)) + 
  geom_vline(xintercept = 0, col = "red", lty = 2) +
  geom_density() + 
  labs(y = "Density", x = "Posterior") +
  facet_wrap(~ ind, ncol = 2) +
  theme_classic(base_size = 24)

```


Estimates of the overlap of posteriors can be used a statistical significance test. For instance we can compare calling activity between sexes during the active stage:

```{r eval = TRUE}

# p value
pbapply::pboptions(type = "none")

pvals <- pbapply::pbreplicate(num.trees, expr =
sum(md$Sol[sample(nrow(md$Sol)), "M:Activo:sexo_consultaH"] - md$Sol[, "H:Activo:sexo_consultaH"] < 0) / nrow(md$Sol)
)

# plot distributions
ggplot(Y[Y$ind %in% c("H:Activo:sexo_consultaH", "M:Activo:sexo_consultaH"), ], aes(x=values)) + geom_density(aes(group = ind, colour = ind, fill = ind), alpha=0.3) + 
      scale_color_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
    scale_fill_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
  labs(y = "Density", x = "Posterior") +
  theme_classic(base_size = 24)

```

Or whether males responde more to females during the reproductive stage:

```{r, eval = FALSE}

# p value
pvals <- pbapply::pbreplicate(num.trees, expr =
sum(md$Sol[sample(nrow(md$Sol)), "M:Activo:sexo_consultaH"] - md$Sol[, "M:Inactivo:sexo_consultaH"] < 0) / nrow(md$Sol)
)

# mean(pvals)

# plot distributions
ggplot(Y[Y$ind %in% c("M:Activo:sexo_consultaM", "M:Inactivo:sexo_consultaM"), ], aes(x=values)) + geom_density(aes(group = ind, colour = ind, fill = ind), alpha=0.3) + 
    scale_color_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
    scale_fill_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
    labs(y = "Density", x = "Posterior") +
  theme_classic(base_size = 24)

```

Whether males respond more often to females during the active period:
```{r}

# p value
pvals <- pbapply::pbreplicate(num.trees, expr =
sum(md$Sol[sample(nrow(md$Sol)), "M:Activo:sexo_consultaH"] - md$Sol[, "M:Activo:sexo_consultaM"] < 0) / nrow(md$Sol)
)

mean(pvals)

# plot distributions
ggplot(Y[Y$ind %in% c("M:Activo:sexo_consultaH", "M:Activo:sexo_consultaM"), ], aes(x=values)) + geom_density(aes(group = ind, colour = ind, fill = ind), alpha=0.3) + 
    scale_color_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
    scale_fill_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
    labs(y = "Density", x = "Posterior") +
  theme_classic(base_size = 24)

```

Whether females respond differentially to sexes:
```{r}

# p value
pvals <- pbapply::pbreplicate(num.trees, expr =
sum(md$Sol[sample(nrow(md$Sol)), "H:Inactivo:sexo_consultaM"] - md$Sol[, "H:Inactivo:sexo_consultaH"] < 0) / nrow(md$Sol)
)

mean(pvals)

# plot distributions
ggplot(Y[Y$ind %in% c("H:Inactivo:sexo_consultaM", "H:Inactivo:sexo_consultaH"), ], aes(x=values)) + geom_density(aes(group = ind, colour = ind, fill = ind), alpha=0.3) + 
  geom_vline(xintercept = 0, lty = 2, col = "red") +
    scale_color_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
    scale_fill_viridis_d(alpha = 0.7, begin = 0.4, end = 0.8) +
    labs(y = "Density", x = "Posterior") +
  theme_classic(base_size = 24)

```

#### Diagnostic plots

```{r diagnostic plots for model, fig.height = 14}

plot(md$Sol)

```

---

## Sex discrimination using random forest

Measure acoustic structure on data from [call ontogeny paper](https://figshare.com/articles/dataset/Supplementary_materials_Ontogeny_of_an_interactive_call-and-response_system_in_Spix_s_disc-winged_bats_PART_1/11651772)

```{r, eval = FALSE}

# read extended selection table downloaded from figshare (ontogeny paper)
est <- readRDS("./data/ext_sel_tab_inquiry.RDS")

# keep adults only
est.adult <- est[est$age.class == "adults", ]

# check data
table(est.adult$sex)

# check number of calls per ID
tab <- table(est.adult$indiv)

# number of indiv with at least 4 calls
sum(tab > 4)

# keep indiv with at least 4 calls
est.adult.4 <- est.adult[est.adult$indiv %in% names(tab[tab > 4]), ]

# check again
table(est.adult.4$indiv, est.adult.4$sex)

# homogenize sampling rate
est.adult.4 <- resample_est(est.adult.4, samp.rate = 375)

######## measure acoustic structure ######
# measure spectrographic parameters
sp <- specan(est.adult.4, wl = 1100)

# check if there are NAs
anyNA(sp)

# add random variable
sp$random <- rnorm(nrow(sp))

# measure cepstral coeffs
cc <- mfcc_stats(est.adult.4)

# bind all acoustic structure parameters together
sp.cc <- cbind(sp, cc[ , -c(1, 2)])

# scale 
sp.cc[, -c(1, 2)] <- scale(sp.cc[, -c(1, 2)])

# add sex to data for random forest
sp.cc$sex <- est.adult.4$sex

write.csv(sp.cc, "./data/processed/acoustic_parameters_for_random_forest.csv", row.names = FALSE)

```

Model using both spectral parameters and MFCC:

```{r, eval = TRUE}

sp.cc <- read.csv("./data/processed/acoustic_parameters_for_random_forest.csv")

# convert to factor
sp.cc$sex <- as.factor(sp.cc$sex)

# run RF model spectral and cepstral parameters
rfm <- ranger(sex ~ ., data = sp.cc[, -c(1, 2)], num.trees = num.trees, importance = "impurity")

# model results (check out-of-bag error)
rfm

# variable importance
imp <- rfm$variable.importance

```

Model using spectral parameters 

```{r, eval = TRUE}

# run RF model only spectral parameters
rfm_sp <- ranger(sex ~ ., data = sp.cc[, -c(1, 2, grep("\\.cc", names(sp.cc)))], num.trees = num.trees, importance = "impurity")

# model results (check out-of-bag error)
rfm_sp


```

Model using MFCC 

```{r, eval = TRUE}
# run RF model only spectral parameters
rfm_cc <- ranger(sex ~ ., data = sp.cc[, grep("\\.cc|sex", names(sp.cc))], num.trees = num.trees, importance = "impurity")

# model results (check out-of-bag error)
rfm_cc

# bind  varaible naes and importance
imp_df <- data.frame(var = names(imp), imp = imp)

# check iimportance
imp_df <- imp_df[order(- imp_df$imp), ]

imp_df[1: which(imp_df$var == "random"), ]

# create null model
# table(est.adult.4$sex)
sp.cc.rand <- sp.cc

# randomize position of sex labels
sp.cc.rand$sex <- sample(sp.cc.rand$sex)

# run null model 
nll_rfm <- ranger(sex ~ ., data = sp.cc.rand[, -c(1, 2)], num.trees = num.trees, importance = "impurity")

nll_rfm$prediction.error

```

### Individual effect and null model 

```{r, eval = FALSE}


#### check effect of individuals using ### 
# run model removing one individual at the time
for(i in unique(est.adult.4$indiv))
jck_nf <- sapply(unique(est.adult.4$indiv), function(i)
ranger(sex ~ ., data = sp.cc[est.adult.4$indiv != i, -c(1, 2)], num.trees = num.trees, importance = "impurity")$prediction.error)

# check prediction errors
jck_nf
range(jck_nf)

sd(jck_nf)

#### check effect of individuals on null model ### 
# run model removing one individual at the time
for(i in unique(est.adult.4$indiv))
jck_nf_null <- sapply(unique(est.adult.4$indiv), function(i)
ranger(sex ~ ., data = sp.cc.rand[est.adult.4$indiv != i, -c(1, 2)], num.trees = num.trees, importance = "impurity")$prediction.error)

# check prediction errors
jck_nf_null
range(jck_nf_null)

jk_res <- data.frame(error = c(jck_nf, jck_nf_null), test = rep(c("real.data", "null.model"), each = length(jck_nf)))

null_errors <- pbapply::pbreplicate(num.trees, expr = ranger(sex ~ ., data = sp.cc.rand[, -c(1, 2)], num.trees = num.trees, importance = "impurity")$prediction.error
)

jk_res <- rbind(jk_res, data.frame(error = null_errors, test = "null_repeated"))

write.csv(jk_res, "./data/processed/jacknifed_random_forest_error.csv", row.names = FALSE)

```

#### jackknife random forest results

 - null.repeated = the null random forest model repeated (without jackknife)
 - null.model = null model with jacknife
 - real.data = real data!
 
```{r, eval = TRUE}

jk_res <- read.csv("./data/processed/jacknifed_random_forest_error.csv")

jk_res$error <- 1 - jk_res$error

agg_jk_res <- aggregate(error ~ test, data = jk_res, FUN = mean)

agg_jk_res$sd <- aggregate(error ~ test, data = jk_res, FUN = sd)$error

ggplot(agg_jk_res, aes(x = test, y = error)) + 
  geom_point(fill = viridis(10)[2]) +
  geom_errorbar(aes(ymin = error - sd, ymax = error + sd), width = 0.1) +  
  labs(x = "Random forest models", y = "Sex discrimination (1 - out-of-bag error") + 
  theme_classic()

```


---

<font size="4">R session information</font>

```{r session info, echo=F}

sessionInfo()

```
